**Machine learning identifies unique taxa differentiating proximal and distal

### 2017-06-12 submission/manuscript.Rmd
refs almost done

Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012). Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_, _Porphyromonas asacharolytica_ and _Parvimonas micra_. Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (@Baxter2016).  However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors. However, the contribution of the gut microbiota to these subtypes is largely undefined.

@@

When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes. This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome. Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (cite). Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (cite). That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (cite). Thus, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.

We identified a range of Simpson diversity values between patients and between each anatomical site. As this measurement takes both number of species and abundance of species into account, it seems that there is a varied level of each component along the human colon. There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal colon, some patients had the opposite. An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (Eckberg 2005). We used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other. We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants. This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (Eckberg 2005, Aguierre de Carcer, 2015, Zhang et al 2013).

To get around the noisiness created by a diverse set of samples, we built a Random Forest model to identify microbes specific to each side. For each comparison we identified top X OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (cite nature review). For instance, the right mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon (cite). The left mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. The model was less effective at classifying the right and left lumenal contents, probably because the samples are arguably composed of the same material.

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (cite Niel) and have oncogenic properties in cell culture and in mice (cite). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (cite something, Albenberg, 2014). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (cite Mima 2016). Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (Mima 2016). In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F. nucleatum_ in mucosal biopsies (Strauss 2011). These studies have the added caveat of being sampled from prepared or cleaned colons. Thus, the presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Indeed, rather than the _F. nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (Flynn 2016). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis.

>>>>

Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012). Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated based on FISH staining and 16S rRNA sequencing (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, the periodontal pathogens are also highly predictive of if a patient has CRC tumors or not in our classification studies (@Baxter2016).  However, these studies have been performed comparing shed stool or in mice, which does not allow for analysis of paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors. However, the contribution of the gut microbiota to these subtypes is largely undefined.

@@

When examining the relative abundance of the different phyla at each site, there is a wide amount of variation for each phyla with communities primarily dominated by the Bacteriodes and Firmicutes. This likely reflects not only the variability between human subjects, casued by differences in age, gender, diet, but also reports of microbial "patchiness" in the gut microbiome. Several previous studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other (@Hong2011). Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance (@Stearns2011). That said, across our samples the mucosal samples harbor more Proteobacteria, consistent with previous studies comparing mucosal swabs to lumenal content in humans (@Albenberg2014). Thus, the conclusions we can draw from phyla analysis are likely confounded by differences in sampling and patchiness between subjects.

We identified a range of Simpson diversity values between patients and between each anatomical site. As this measurement takes both number of species and abundance of species into account, it seems that there is a varied level of each component along the human colon. There was no clear pattern of diversity when comparing patients- some patients had a higher level of diversity in the proximal than the distal colon, some patients had the opposite. An early study of microbial diversity and membership of three human subjects found a varied level of diversity in each subject, but that subjects were overall more similar to themselves than to each other (@Eckberg2005). We used the thetaYC dissimilarity metric to compare the communities present at each anatomical location to each other. We found that while there was a wide range of similarity of each site within a person, the study participants were ultimately more similar to themselves than all of the other participants. This finding is consistent with numerous studies examining the biogeography of the colon in healthy humans (@Eckberg2005, @deCarcer2010, @Zhang2013).

To get around the noisiness created by a diverse set of samples, we built a Random Forest model to identify microbes specific to each side. For each comparison we identified top X OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis (@Donaldson2015). For instance, the right mucosa harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and the aerobic _Psuedomonas_ consistent with the highest oxygen regions of the colon. The left mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. The model was less effective at classifying the right and left lumenal contents, probably because the samples are arguably composed of the same material.

We detected _F. nucleatum_ and _P. asacharolytica_ in 7 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans (@Baxter2016) and have oncogenic properties in cell culture and in mice (@Sears2014). Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the left mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous left mucosa (@Albenberg2014). In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on right-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum (@Mima2016). Of the (8) (40%) individuals positive for _F. nucleatum_ in this study, the bacteria was spread across the right mucosa, left lumen and left mucosa. That said, in a previous study of 1,102 colorectal cancer cases, only 138 (11%) of them contained detectable _F. nucleatum_ at all (@Mima2016). In a separate study of patients with inflammatory bowel disease (IBD), 50% of 22 IBD patients and 17.6% of 34 people without disease contained _F. nucleatum_ in mucosal biopsies (@Strauss2011). These studies have the added caveat of being sampled from prepared or cleaned colons. Thus, the presence of _F. nucleatum_ in a healthy individual is not necessarily linked to the development of future colorectal cancers. Because of the spatial distribution of the _F. nucleatum_ in our sample set, we cannot develop a model for the role of _F. nucleatum_ in the healthy colon. Indeed, rather than the _F. nucleatum_ initiating tumorigenesis, these bacteria may instead arrive "late" to the tumor microenvironment and feed on tumors that have alread begun to develop (@Flynn2016). Data examining bacterial biofilms on CRC tumors suggests that _Fusobacteria_ species are more commonly found both on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis. Finally, _Fusobacteria_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model (@Whitmore2014, @Flynn2016). Thus, further analysis of the distribution and role of these pathogens may elucidate a mechanism for development of CRC tumors in the proximal or distal colon.
